Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-05-07 LICS2021-final-for-proceedings.tex

%%line and pagenumbers to be removed for final version
%\usepackage[switch]{lineno}
\pagestyle{plain}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\IEEEoverridecommandlockouts
\IEEEpubid{\makebox[\columnwidth]{\emph{[Long version of the LICS 2021 paper, with full proofs in the appendix.]}\hfill}
    \hspace{\columnsep}\makebox[\columnwidth]{ }}
%\IEEEpubid{\makebox[\columnwidth]{978-1-6654-4895-6/21/\$31.00~
%\copyright2021 IEEE \hfill} \hspace{\columnsep}\makebox[\columnwidth]{ }}

@@

\begin{thm}[\cite{duffin:65}]\label{thm:cograph}
  A graph is a cograph iff it can be constructed from the singletons
  via the operations $+$ and~$\times$.
\end{thm}

@@

\begin{proof}
  First, we can replace every instance of the $\fff$-rule in $\Deri$ by
  $\weakr$. Then the instances of $\weakr$ and $\conr$ are replaced by
  $\wD$ and $\cD$, which can then be permuted down.
  Details are in
  Appendix~\ref{app:LK1-decompose}.
\end{proof}

@@

\begin{proof}
  Permute all $\wD$ and $\wfaD$ instances to the bottom of
  the derivation, then permute all $\cD$ and
  $\cfaD$ below $\set{\me,\mfaD,\mexD}$. This involves a tedious but
  routine case analysis. However, unlike most other rule
  permutations in this paper, this has not been done before in the deep
  inference literature. 
  For this reason, we give the full case
  analysis in Appendix~\ref{app:cw-decomposition}.
  This
  Lemma is the reason for the presence of the rules $\wfaD$ and
  $\cfaD$, as without them the permutation cases in~\eqref{eq:wfa-cfa}
  could not be resolved.
\end{proof}

@@

\begin{proof}
  By \cite[Proposition~7.5]{str:ral:tableaux19}, there is a derivation
  $\vlder{\set{\acD,\me,\PE\fequ}}{\DDeri}{\PE
    H}{\PE{(G\substof\phi)}}$.  We can lift $\DDeri$ to a first-order
  derivation in $\set{\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$, in a similar
  way as in the previous lemma.
  The technical details are in
  Appendix~\ref{app:surj->cm}.
\end{proof}

@@

%\end{document}

%\clearpage
\newpage
\appendix

\subsection{Proof of Theorem~\ref{thm:LK1-decompose}}
\label{app:LK1-decompose}

\begin{proof}[Proof of Theorem~\ref{thm:LK1-decompose}]

Write $\fv(A)$ for the set of variables which occur free in $A$.

Note that the instances of $\wrD,\cD$ in $\Deri_2$ are deep, but inside sequent contexts.

First, if an instance of $\vlinf{\weakr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma}}$
 is followed by a rule in which $A$ is not in the principal
formula, it can be permuted downwards.
Otherwise, the proof can be transformed using the following rewriting rules.

\begin{equation*}
\vlderivation{
  \vliin{\vlan}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}{
    \vlhy{\sqns{B, \Delta}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\vlvdots}{
      \vlhy{\sqns{\Gamma}}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\vlor}{}{\sqns{\Gamma, A \vlor B}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, B}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\sqns{\Gamma, A \vlor B}}{
    \vlhy{\sqns{\Gamma, A}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\exists}{}{\sqns{\Gamma, \exists x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A\ssubst{x}{t}}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, \exists x.A}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\forall}{}{\sqns{\Gamma, \forall x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, \forall x.A}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqns{\Gamma, A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, A}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlhy{\sqns{\Gamma, A}}}
\end{equation*}
Note that in the case of $\vlor$, we use the deep rule $\wrD$ which can be
permuted under all the rules. By using these rewriting rules, we can eventually get
a derivation with all the instances of $\weakr$ and $\wrD$ at the bottom. Now observe
that the instances of $\conr$ in $\Deri$ can be transformed using the following
rule:

\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqns{\Gamma, A}}{
    \vlhy{\sqns{\Gamma, A, A}}}}
\leadsto
\vlderivation{
  \vlin{\cD}{}{\sqns{\Gamma, A}}{
    \vlin{\vlor}{}{\sqns{\Gamma, A \vlor A}}{
      \vlhy{\sqns{\Gamma, A, A}}}}}
\end{equation*}

Knowing that $\cD$ can be permuted under all the rules of $\FOMLL$, we
eventually obtain a derivation:
\begin{equation*}
\vlderivation{
  \vlde{\set{\weakr, \wrD,\cD,\fequ}}{\Deri'_2}{\sqns{\Gamma}}{
    \vltrl{}{\FOMLL}{\Deri'_1}{\sqns{\Gamma'}}{
      \vlhy{\quad}}{
      \vlhy{}}{
      \vlhy{}}}}
\end{equation*}
Note that $\fequ$ is required here since the permutation of formulas is implicit
in $\FOMLL$.

By transforming each sequent of $\Deri'_2$ into its corresponding formula, and by
considering the following rewriting rule:
\begin{equation*}
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A}}{
    \vlhy{\sqns{\Gamma}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\sqns{\form{\Gamma} \vlor A}}{
    \vlhy{\sqns{\form{\Gamma}}}}}
\end{equation*},
we obtain a derivation
\begin{equation*}
\vlderivation{
  \vlde{\set{\wrD,\cD,\fequ}}{\Deri_2}{\sqns{\form{\Gamma}}}{
    \vltrl{}{\FOMLL}{\Deri_1}{\sqns{\form{\Gamma'}}}{
      \vlhy{\quad}}{
      \vlhy{}}{
      \vlhy{}}}}
\end{equation*}
where $\Deri_1$ can be obtained from $\Deri'_1$
by applying the $\vlor$ rule.
\end{proof}

\subsection{Rule permutation for the proof of Lemma~\ref{lem:cw-decomposition}}
\label{app:cw-decomposition}
We construct a rewriting system based on rule permutation on derivations in $\set{\wrD, \wfaD, \acD,
\cfaD, \me, \mfaD, \mexD, \fequ}$ that allows us to reach a derivation of the
form 
\begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\fequ}}{}{B}{
        \vlde{\set{\acD,\cfaD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
\end{equation*}
from any derivation. Intuitively, we want to move all the instances of $\rr \in
\set{\wrD, \wfaD}$ downwards and all the instances of $\rr' \in \set{\me,
\mfaD, \mexD}$ upwards.

We first study the interactions between two rules. Certain cases are unsolved at
this stage, and they are considered later when we study the interactions between
two non-$\fequ$ rule instances separated by $\fequ$. Only non-trivial cases are
presented here:
\begin{itemize}
\item $\rr_1/\rr_2$, where $\rr_1 \in \set{\wrD, \wfaD}$ and $\rr_2 \in
\set{\acD, \cfaD, \me, \mfaD, \mexD}$:

\begin{equation*}
\vlderivation{
  \vlin{\acD}{}{a}{
    \vlin{\wrD}{}{a \vlor a}{ 
      \vlhy{a}}}}
\leadsto
\vlderivation{
  \vlhy{a}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\me}{}{(A \vlor B) \vlan (C \vlor D)}{
    \vlin{\wrD}{}{(A \vlan C) \vlor (B \vlan D)}{
      \vlhy{A \vlan C}}}}
\leadsto 
\vlderivation{
  \vlin{\wrD}{}{(A \vlor B) \vlan (C \vlor D)}{
    \vlin{\wrD}{}{(A \vlor B) \vlan C}{
      \vlhy{A \vlan C}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\wrD}{}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\forall x.(A \vlor B)}{
    \vlhy{\forall x.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall x.A}{
    \vlin{\wfaD}{}{\forall x.\forall x.A}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlhy{\forall x.A}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{A \vlor (\forall x.B)}}}}
\leadsto
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor (\forall x.B)}}} 
\end{equation*}
where in the last case, $x$ is not free in $A$.

\item $\rr_1/\rr_2$, where $\rr_1 \in \set{\acD, \cfaD}$ and $\rr_2 \in
\set{\me, \mfaD, \mexD}$: 
  \begin{equation*}
  \scalebox{.9}{$
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\cfaD}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mfaD}{}{\Scons{\forall x.(A \vlor B)}}{
      \vlin{\fequ}{}{\Scons{(\forall x.A) \vlor (\forall x.B)}}{
        \vlin{\mfaD}{}{\Scons{\forall x.(\forall x.A \vlor B)}}{
          \vlhy{\Scons{(\forall x.\forall x.A) \vlor (\forall x.B)}}}}}}$}
  \end{equation*}

\item $\cfaD/\fequ$:

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall y.\forall x.A}{
    \vlin{\cfaD}{}{\forall x.\forall y.A}{
      \vlhy{\forall x.\forall x.\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\cfaD}{}{\forall y.\forall x.A}{
    \vlin{\fequ}{}{\forall y.\forall x.\forall x.A}{
      \vlhy{\forall x.\forall x.\forall y.A}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlhy{\forall x.\forall x.(A \vlor B)}}}}
\leadsto
\vlderivation{
  \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
    \vlin{\fequ}{}{(\forall x.\forall x. A) \vlor B}{
      \vlhy{\forall x.\forall x.(A \vlor B)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlhy{(\forall x.\forall x.A) \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\fequ}{}{\forall x.\forall x.(A \vlor B)}{
      \vlhy{(\forall x.\forall x.A) \vlor B}}}}
\end{equation*}
where in the last two cases, $x$ is not free in $B$.

\item $\wrD/\fequ$:

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{B \vlor A}{
    \vlin{\wrD}{}{A \vlor B}{
      \vlhy{A}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{A \vlor (B \vlor C)}{
    \vlin{\wrD}{}{(A \vlor B) \vlor C)}{
      \vlhy{A \vlor C}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{(\forall x. A) \vlor B}{
    \vlhy{\forall x.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\wrD}{}{\forall x.(B \vlor A)}{
      \vlhy{\forall x.B}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\wrD}{}{(\forall x.A) \vlor B}{
      \vlhy{\forall x.A}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\forall x.(A \vlor B)}{
    \vlhy{\forall x.A}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\wrD}{}{B \vlor (\forall x.A)}{
      \vlhy{B}}}}
\end{equation*}
where in the last four cases, $x$ is not free in $B$.
\item $\wfaD/\fequ$:

In the following two cases, we assume $x \neq y$ (otherwise they are trivial).

\begin{equation*}
\scalebox{.9}{$
\vlderivation{
  \vlin{\fequ}{}{\forall y.\forall x.A}{
    \vlin{\wfaD}{(x \notin \fv(\forall y.A))}{\forall x.\forall y.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin \fv(A))}{\forall y.\forall x.A}{
    \vlhy{\forall y.A}}}$}
\end{equation*}

\begin{equation*}
\scalebox{.9}{$
\vlderivation{
  \vlin{\fequ}{}{\forall x.\forall y.A}{
    \vlin{\wfaD}{(x \notin \fv(A))}{\forall y.\forall x.A}{
      \vlhy{\forall y.A}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{(x \notin \fv(\forall y.A))}{\forall x.\forall y.A}{
    \vlhy{\forall y.A}}}$}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{(\forall x.A) \vlor B}{
    \vlin{\wfaD}{}{\forall x.(A \vlor B)}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{}{(\forall x.A) \vlor B}{
    \vlhy{A \vlor B}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\wfaD}{}{(\forall x.A) \vlor B}{
      \vlhy{A \vlor B}}}}
\leadsto
\vlderivation{
  \vlin{\wfaD}{}{\forall x.(A \vlor B)}{
    \vlhy{A \vlor B}}}
\end{equation*}
where in the last two cases, the constraint on $x$ on the left-hand side implies
that of the right-hand side.

\item $\fequ/\cfaD$:

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall x.\forall y.A}{
    \vlin{\fequ}{}{\forall x.\forall x.\forall y. A}{
      \vlhy{\forall x.\forall y.\forall x.A}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall y.\forall x.A}{
    \vlin{\fequ}{}{\forall y.\forall x.\forall x.A}{
      \vlhy{\forall x.\forall y.\forall x.A}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
    \vlin{\fequ}{(x \notin \fv(B))}{(\forall x.\forall x.A) \vlor B}{
      \vlhy{\forall x.((\forall x.A) \vlor B)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\fequ}{(x \notin \fv(B))}{\forall x.\forall x.(A \vlor B)}{
      \vlhy{\forall x.((\forall x.A) \vlor B)}}}}
\end{equation*}

\item $\fequ/\me$:

\begin{equation*}
\vlderivation{
  \vlin{\me}{}{(A \vlor B) \vlan (C \vlor D)}{
    \vlin{\fequ}{}{(A \vlan C) \vlor (B \vlan D)}{
      \vlhy{(C \vlan A) \vlor (B \vlan D)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\me}{}{(A \vlor B) \vlan (C \vlor D)}{
    \vlin{\fequ}{}{(A \vlan C) \vlor (B \vlan D)}{
      \vlhy{(B \vlan D) \vlor (A \vlan C)}}}}
\leadsto
\vlderivation{
  \vlin{\fequ}{}{(A \vlor B) \vlan (C \vlor D)}{
    \vlin{\me}{}{(B \vlor A) \vlan (D \vlor C)}{
      \vlhy{(B \vlan D) \vlor (A \vlan C)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\me}{}{(A \vlor B) \vlan ((C \vlan E) \vlor D)}{
    \vlin{\fequ}{}{(A \vlan (C \vlan E)) \vlor (B \vlan D)}{
      \vlhy{((A \vlan C) \vlan E) \vlor (B \vlan D)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\me}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
    \vlin{\fequ}{(x \notin \fv(B \vlan D))}{\forall x.((A \vlan C) \vlor (B \vlan D))}{
      \vlhy{(\forall x.(A \vlan C)) \vlor (B \vlan D)}}}}
\end{equation*}

\item $\fequ/\mfaD$:

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\fequ}{}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{(\forall x.B) \vlor (\forall x.A)}}}}
\leadsto
\vlderivation{
  \vlin{\fequ}{}{\forall x.(A \vlor B)}{
    \vlin{\mfaD}{}{\forall x.(B \vlor A)}{
      \vlhy{(\forall x.B) \vlor (\forall x.A)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.((\forall y.A) \vlor B)}{
    \vlin{\fequ}{}{(\forall x.\forall y.A) \vlor (\forall x.B)}{
      \vlhy{(\forall y.\forall x.A) \vlor (\forall x.B)}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
    \vlin{\fequ}{}{(\forall x.A) \vlor (\forall x.B)}{
      \vlhy{\forall x.(A \vlor (\forall x.B))}}}}
\end{equation*}

\item $\fequ/\mexD$: similar to $\fequ/\mfaD$

\end{itemize}

Interactions between two non-$\fequ$ rules with the presence of $\fequ$ in
between:

\begin{itemize}
\item $\cfaD/\fequ/\rr$ where $\rr \in \set{\me, \mfaD, \mexD}$: First permute
$\cfaD$ under $\fequ$ and then permute $\cfaD$ under $\rr$.

\item $\acD/\fequ/\rr$ where $\rr \in \set{\me, \mfaD, \mexD}$: First permute
$\acD$ under $\fequ$ and then permute $\acD$ under $\rr$.

\item $\wrD/\fequ/\cfaD$:
  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlin{\fequ}{}{(\forall x.\forall x.A) \vlor B}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{(\forall x.A) \vlor B}{
      \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}
  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{(\forall x.A) \vlor B}{
      \vlin{\fequ}{}{(\forall x.\forall x.A) \vlor B}{
        \vlin{\wrD}{}{\forall x.(B \vlor (\forall x.A))}{
          \vlhy{\forall x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{(\forall x.A) \vlor B}{
      \vlin{\wrD}{}{\forall x.(B \vlor A)}{
        \vlhy{\forall x.B}}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{\forall x.\forall x.(A \vlor B)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlin{\cfaD}{}{\forall x.A}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\cfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{\forall x.\forall x.(A \vlor B)}{
        \vlin{\wrD}{}{\forall x.(B \vlor (\forall x.A))}{
          \vlhy{\forall x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.(A \vlor B)}{
      \vlin{\wrD}{}{\forall x.(B \vlor A)}{
        \vlhy{\forall x.B}}}}
  \end{equation*}
  where in all four cases, $x$ is not free in $B$.
\item $\wrD/\fequ/\acD$:
  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{a \vlor B}{
      \vlin{\fequ}{}{(a \vlor a) \vlor B}{
        \vlin{\wrD}{}{(a \vlor B) \vlor a}{
          \vlhy{a \vlor B}}}}}
  \leadsto
  \vlderivation{
    \vlhy{a \vlor B}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{a \vlor B}{
      \vlin{\fequ}{}{(a \vlor a) \vlor B}{
        \vlin{\wrD}{}{a \vlor (a \vlor B)}{
          \vlhy{a}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{a \vlor B}{
      \vlhy{a}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{\forall x.a}{
      \vlin{\fequ}{(x \notin \fv(a))}{\forall x.(a \vlor a)}{
        \vlin{\wrD}{}{(\forall x.a) \vlor a}{
          \vlhy{\forall x.a}}}}}
  \leadsto
  \vlderivation{
    \vlhy{\forall x.a}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\acD}{}{\forall x.a}{
      \vlin{\fequ}{(x \notin \fv(a))}{\forall x.(a \vlor a)}{
        \vlin{\wrD}{}{a \vlor (\forall x.a)}{
          \vlhy{a}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wfaD}{(x \notin \fv(a))}{\forall x.a}{
      \vlhy{a}}}
  \end{equation*}

\item $\wrD/\fequ/\me$:
  \begin{equation*}
  \vlderivation{
    \vlin{\me}{}{(A \vlor B) \vlan (C \vlor D)}{
      \vlin{\fequ}{}{(A \vlan C) \vlor (B \vlan D)}{
        \vlin{\wrD}{}{(C \vlan A) \vlor (B \vlan D)}{
          \vlhy{C \vlan A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{(A \vlor B) \vlan (C \vlor D)}{
      \vlin{\wrD}{}{(A \vlor B) \vlan C}{
        \vlin{\fequ}{}{A \vlan C}{
          \vlhy{C \vlan A}}}}}
  \end{equation*}

  \begin{equation*}
  \scalebox{.9}{$
  \vlderivation{
    \vlin{\me}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\fequ}{}{\forall x.((A \vlan C) \vlor (B \vlan
D))}{
        \vlin{\wrD}{}{(B \vlan D) \vlor (\forall x.(A \vlan C))}{
          \vlhy{B \vlan D}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.((A \vlor B) \vlan (C \vlor D))}{
      \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan (D \vlor C))}{
        \vlin{\wrD}{}{\forall x.((B \vlor A) \vlan D)}{
          \vlin{\faD}{}{\forall x.(B \vlan D)}{
            \vlhy{B \vlan D}}}}}}$}
  \end{equation*}
  where in the second case, $x$ is free in $B \vlan D$.

\item $\wrD/\fequ/\mfaD$:
  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{(\forall x.A) \vlor (\forall x.B)}{
        \vlin{\wrD}{}{(\forall x.B) \vlor (\forall x.A)}{
          \vlhy{\forall x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\forall x.(A \vlor B)}{
      \vlin{\wrD}{}{\forall x.(B \vlor A)}{
        \vlhy{\forall x.B}}}}
  \end{equation*}

  \begin{equation*}
  \vlderivation{
    \vlin{\mfaD}{}{\forall x.(A \vlor B)}{
      \vlin{\fequ}{}{(\forall x.A) \vlor (\forall x.B)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor B)}{
          \vlhy{\forall x.\forall x.A}}}}}
  \leadsto
  \vlderivation{
    \vlin{\wrD}{}{\forall x.(A \vlor B)}{
      \vlin{\cfaD}{}{\forall x.A}{
        \vlhy{\forall x.\forall x.A}}}}
  \end{equation*}

\item $\wrD/\fequ/\mexD$:

  \begin{equation*}
  \vlderivation{
    \vlin{\mexD}{}{\exists x.(A \vlor B)}{
      \vlin{\fequ}{}{(\exists x.A) \vlor (\exists x.B)}{
        \vlin{\wrD}{}{(\exists x.B) \vlor (\exists x.A)}{
          \vlhy{\exists x.B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\fequ}{}{\exists x.(A \vlor B)}{
      \vlin{\wrD}{}{\exists x.(B \vlor A)}{
        \vlhy{\exists x.B}}}}
  \end{equation*}

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Proof of Lemma~\ref{lem:surj->cm}}
\label{app:surj->cm}

\begin{proof}[Proof of Lemma~\ref{lem:surj->cm}]
  By
  \cite[Proposition~7.5]{str:ral:tableaux19}, there is a derivation
  $\vlder{\set{\acD,\me,\PE\fequ}}{\DDeri}{\PE H}{\PE{(G\rsubstof\phi)}}$,
   We plan to show that $\DDeri$ can be lifted to
  $\set{\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$. However, observe that not
  every formula occurring in $\DDeri$ is a propositional
  encoding. There are two reasons for this: (i) we might have
  $P\PE\fequ Q$ where $P$ is a propositional encoding but $Q$ is not,
  and (ii) the rule $\acD$ can duplicate an atom
  $x\in\VAR$. Let us write $\acDx$ for such instances.
  The problem with (i) is that we could have the following situation
  \begin{equation}
    \label{eq:m-illegal}
      \vlin{\me}{}{\Scons{((x\vlan E)\vlor(x\vlan F))\vlan(C\vlor D)}}{
        \vlin{\PE\fequ}{}{\Scons{((x\vlan E)\vlan C)\vlor((x\vlan F)\vlan D)}}{
          \vlhy{\Scons{(x\vlan (E\vlan C))\vlor(x\vlan (F\vlan D))}}}}}
  \end{equation}
  where $x$ occurs in $C\vlor D$. Then premise and conclusion are both
  propositional encodings, but the whole derivation cannot be
  lifted. However, since we demand that the mapping is a fibration
  (and therefore a homomorphism) on the binding graphs, there must be
  another instance of $\me$ further below in the derivation:
  \vspace{-2ex}\begin{equation}
      \vlin{\me}{}{S'\cons{(x\vlor x)\vlan(E\vlor F)}}{
        \vlhy{S'\cons{(x\vlan E)\vlor(x\vlan F)}}}}
  \end{equation}
  We can permute both instances via the following more general scheme
  (see~\cite{str:07:RTA,lamarche:gap} for a general discussion on
  permutations of the $\me$-rule):
  \begin{equation}
    \label{eq:m-permutation}
    \scalebox{.66}{$
      \vlin{\me}{}{\Scons{(G\vlor H)\vlan(E\vlor F)\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{((G\vlan E)\vlor(H\vlan F))\vlan(C\vlor D)}}{
          \vlhy{\Scons{(G\vlan E\vlan C)\vlor(G\vlan F\vlan D)}}}}}
    \;
    \leftrightarrow
    \;
      \vlin{\me}{}{\Scons{(G\vlor H)\vlan(E\vlor F)\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{(G\vlor H)\vlan((E\vlan C)\vlor(F\vlan D))}}{
          \vlhy{\Scons{(G\vlan E\vlan C)\vlor(G\vlan F\vlan D)}}}}}    
    $}
  \end{equation}
  We omitted some instances of $\PE\fequ$ and some parentheses.  We
  now call instances of $\me$ as in~\eqref{eq:m-illegal}
  \bfit{illegal}, and we can transform $\DDeri$ through
  $\me$-permutations~\eqref{eq:m-permutation} into a derivation that
  does not contain any illegal $\me$-instances.
  To
  address (ii),
  we also apply a permutation argument, permuting all
  instances of $\acDx$ up until they either reach the top of the
  derivation or an instance of $\me$ which separates the two atoms in
  the premise. More precisely, we consider the following inference rule
  \begin{equation}
    \label{eq:acDeq}
    \vlinf{\acDeq}{}{\Scons{x}}{S_0\cons{S_1\cons{x}\vlor S_2\cons{x}}}
  \end{equation}

>>>>

%%line and pagenumbers to be removed for final version
%\usepackage[switch]{lineno}
%\pagestyle{plain}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\IEEEoverridecommandlockouts
\IEEEpubid{\makebox[\columnwidth]{978-1-6654-4895-6/21/\$31.00~
\copyright2021 IEEE \hfill} \hspace{\columnsep}\makebox[\columnwidth]{ }}

@@

\begin{thm}[\cite{Ler71,CLS81}]\label{thm:cograph}
  A graph is a cograph iff it can be constructed from the singletons
  via the operations $+$ and~$\times$.
\end{thm}

@@

\begin{proof}
  First, we can replace every instance of the $\fff$-rule in $\Deri$ by
  $\weakr$. Then the instances of $\weakr$ and $\conr$ are replaced by
  $\wD$ and $\cD$, which can then be permuted down.
  Details are in
  Appendix~A of~\cite{HSW21:arxiv}.
\end{proof}

@@

\begin{proof}
  Permute all $\wD$ and $\wfaD$ instances to the bottom of
  the derivation, then permute all $\cD$ and
  $\cfaD$ below $\set{\me,\mfaD,\mexD}$. This involves a tedious but
  routine case analysis. However, unlike most other rule
  permutations in this paper, this has not been done before in the deep
  inference literature. 
  For this reason, we give the full case
  analysis in Appendix~B of~\cite{HSW21:arxiv}.
  This
  Lemma is the reason for the presence of the rules $\wfaD$ and
  $\cfaD$, as without them the permutation cases in~\eqref{eq:wfa-cfa}
  could not be resolved.
\end{proof}

@@

\begin{proof}
  By \cite[Proposition~7.5]{str:ral:tableaux19}, there is a derivation
  $\vlder{\set{\acD,\me,\PE\fequ}}{\DDeri}{\PE
    H}{\PE{(G\substof\phi)}}$.  We can lift $\DDeri$ to a first-order
  derivation in $\set{\acD,\cfaD,\me,\mfaD,\mexD,\fequ}$, in a similar
  way as in the previous lemma.
  The technical details are in
  Appendix~C of~\cite{HSW21:arxiv}.
\end{proof}

@@
